Derogations and Democratic Backsliding: Exploring the Pandemic's Effects on Civic Spaces

### 2021-03-20 manuscript/manuscript.Rmd
Major updates and edits to manuscript

---
title: "Derogations and Democratic Backsliding: Exploring the Pandemic's Effects on Civic Spaces"
author:
- name: Suparna Chaudhry
  affiliation: Lewis & Clark College
  email: schaudhry@lclark.edu
  url: http://www.suparnachaudhry.com/
- name: Andrew Heiss
  affiliation: Georgia State University
  email: aheiss@gsu.edu
  url: https://www.andrewheiss.com/
date: March 21, 2021
published: This is an early draft of a working paper.
git-repo: https://example.com
abstract: >-
  In an effort to combat the COVID-19 pandemic, most governments have imposed restrictions on movement, association, and other civic freedoms in the interest of public health. In many cases, such as New Zealand, these emergency measures have been temporary and respect for human rights has returned to normal. In many other instances, however, governments have used these restrictions to suppress opposition and more permanently restrict civic space. Systematically measuring the consequences of COVID restrictions, however, is a difficult task. We will examine two possible quantitative measures of the relationship between of COVID restrictions and civil society space. First, we will use the Variety of Democracy project's newly released Pandemic Violations of Democratic Standards Index to explore if and how civil society restrictions predict pandemic backsliding. Second, while many countries sign international human rights treaties that ostensibly bind states to respect rights, several treaties allow for emergency derogations from these obligations. We will tabulate formal human rights treaty derogations due to the pandemic and explore whether these emergency measures led to lasting declines in associational and human rights. We hope that our exploration of these two measures will provide rich data-based descriptions of the relationship between COVID restrictions and civic space that will allow for more causal work in the future.
reference-section-title: References
---

In an effort to combat the COVID-19 pandemic, governments across the world have imposed restrictions on freedom of movement, association, assembly and other civic freedoms in the interest of public health. Many of these states have done so by declaring states of emergency and formally derogating from their international treaty obligations. Yet, far more countries have adopted emergency measures and restricted individual liberties than have derogated from human rights treaties. Even within states that have formally derogated, some states' responses have been temporary and proportionate as advised by the United Nations Office of the High Commissioner of Human Rights [source: https://www.ohchr.org/Documents/Events/EmergencyMeasures_COVID19.pdf] - while others have used these measures to suppress opposition and more permanently restrict civic space. During the COVID-19 pandemic, how many countries have formally derogated from their human rights treaty obligations and how does the nature of measures implemented during this period differ across countries? In the long-term, we aim to use our rich data-based description of the relationship between pandemic restrictions and civic space to examine whether these emergency measures led to lasting declines in associational and human rights, furthering the phenomenon of closing civic space.

Studying state compliance with the international human rights treaty obligations as well as the (unintended) effects of derogations is important, especially given the ongoing prior phenomenon of closing civic space. Even prior to the pandemic, since 2013, 91 countries had proposed or enacted 244 measures restricting civil society [source: https://www.icnl.org/our-work/global-programs/the-civic-space-initiative].  In Human Rights Watch’s 2016 World Report, Executive Director Kenneth Roth argued that civil society was under more aggressive attack than at any time in recent memory. ANDREW: DO YOU WANT TO INCLUDE SOME SUMMARY FIGURES FROM OUR DATA/CLOSING SPACE CHAPTER??

Once the pandemic began, according to the International Center for Not-for-Profit Law (ICNL) 107 countries declared emergency, 56 implemented measures affecting freedom of expression and even more -- 139 countries implemented measures affecting freedom of assembly [source: https://www.icnl.org/covid19tracker/]. Many others have erected new barriers to public information, prevented NGOs from delivering services. In seeking to target disinformation, many countries have penalized dissent and securitized the response to the pandemic, using it as a justification for reducing civic freedoms. Other governments have used pandemic restrictions as an excuse for delaying elections [https://www.ifes.org/publications/global-impact-covid-19-elections] or for denying arrested individuals adequate legal representation in judicial proceedings [https://www.unodc.org/documents/Advocacy-Section/Ensuring_Access_to_Justice_in_the_Context_of_COVID-191.pdf].

While some countries have implemented these measures without formally derogating from international human rights treaties, otherwise have esssentially violated their treaty commitments. But even among countries that formally derogate, it is worth documenting measures as many countries have renewed their derogations -- leading to concerns about serial derogators and its potential consequences on measures that are neither proportional nor temporary.

Using original data on derogations to the International Covenant on Civil and Political Rights (ICCPR) during the pandemic as well as data on measures implemented during Covid-19 from PanDem by V-DEM, we find 1) HOW MANY COUNTRIES ACTUALLY DEROGATED 2) Countries that implement emergency measures while legally derogating from their treaty commitments ARE DOING WHAT/CONFIRM less likely to violate non-derogable rights during the pandemic ONCE WE HAVE FULL DATA 3) ANY OTHER FINDINGS

# Derogations

Derogations in international law authorize states to temporarily escape their international treaty commitments during times of crises (Hafner-Burton et al 2011, 647). These derogations are especially important from a human rights perspective, since they are triggered during crises, when individual liberies are under serious threat (677). By declaring a state of public emergency and formally derogating, states acknowledge that these measures are temporary, necessary, and proportional to the interest at stake, with an aim to restore normalcy as soon as possible. Without this option, states facing emergencies may outrightly violate their treaty commitments under international law. Further, they may also illegally violate non-derogable rights such as right to life, freedom from torture and slavery.

While the most frequent derogators are stable democracies and countries where domestic courts can exercise strong oversight of the executive and hold them responsible for breaches of human rights agreements, derogations, especially in the context of a long-lasting event such as the pandemic are more concerning because of the tendency of some countries to turn into ``serial derogators." These countries generally derogate ``without providing information about rights restrictions and in multiple consecutive years." (Hafner-Burton et al 675) These patterns are typically encountered in countries where the judiciary is weak and voters cannot easily remove leaders from office.

@@

The exceptional circumstances brought on by the global pandemic has led to numerous extensive derogations. International human rights treaties typically lay out circumstances during which derogations may take place -- for the European Convention on Human Rights, derogations can apply `in terms of war or other public emergency threatening the life of the nation,' for the American Convention on Human Rights, a `time of war, public danger, or other emergency that threatens the independence or security of a state.' [source:https://www.ohchr.org/Documents/Publications/training9chapter16en.pdf] In the current version of the paper, however, we focus on derogations (and violations) from the International Convenant on Civil and Political Rights, arguably among the most significant and extensive treaties in the international human rights regime.

The ICCPR was adopted by the United Nations’ (UN) General Assembly on December 19, 1966 and came into force on March 23, 1976. Its purpose was to “recognize the inherent dignity of each individual and [to] undertake to promote conditions within states to allow the enjoyment of civil and political rights.” The ICCPR allows the suspension or derogation of certain civil and political rights only under specific situations of emergency that "threaten the life of the nation." [source: https://www.ohchr.org/en/professionalinterest/pages/ccpr.aspx]

Due to the number of states using a variety of measures to curb civil and political rights, the UN Human Rights Office of the High Commissioner published guidance on emergency measures on 27 April 2020, noting that emergency measures must meet the criteria of legality, necessity, proportionality, and non-discrimination. Further, states of emergency are specifically regulated under human rights law and that "states should take measures to prevent human rights violations and abuses associated with the state of emergency perpetrated by state and non-state actors." [source: https://www.ohchr.org/Documents/Events/EmergencyMeasures_COVID19.pdf] It concluded with a reminder of the necessity of democratic practices -- including transparency, right to information, media freedom, rule of law and allowing dissent.

However, since these above democratic freedoms and practices have been under assault owing to closing civic space and more generally, trends in democratic backsliding, it raises concerns that many states may be using emergency measures intended to deal with the pandemic as a way to permanently suppress civic space.

# Data

## Descriptive

# Exploratory case analysis
As shown above, numerous countries that ratified the ICCPR derogated during the pandemic -- many of these countries implemented measures going far beyond the recommended legal, necessary, temporary and proportional nature. These resulting measures, and in the case of countries violating their international treaty commitments without appropriate derogations, there can be numerous consequences for civic space. In the long-term, while we assess this above relationship through the use of systematic data through quantitative means, below, we conduct some exploratory case illustrations. These cases show the range in variation in how states have responded to their international legal commitments - some have derogated as intended, some have officially derogated from their treaty commitments but their measures are arguably not considered as temporary and proportional as necessary. Yet, other countries have avoid derogating completely, thus violating their treaty commitments.

## Derogated as intended

## Officially derogated but measures not temporary and proportional
Armenia reported its first COVID-19 case on March 1st, 2020. As the number of cases rose, the government proclaimed a state of emergency on March 16 of the same year citing Article 3 of the Armenian Constitution which requires “the State to take measures for the protection of the life and health of individuals” [Government of the Republic of Armenia]. Subsequently, Armenia also officially derogated from Article 9, 12 and 21 of the ICCPR, as well as Article 11 of the ECHR March 20, 2020 [FOOTNOTE WHAT THESE ARE]. For both derogations, Armenia extended its original 30 day state of emergency expected to end on April 14th, 2020 another 30 days to May 14th, 2020. They issued five total 30 day extensions on the state of emergency and subsequent derogations. Armenia’s state of emergency finally concluded on September 11th, 2020. It issued a notice of withdrawal of derogation from the ECHR on September 16th, 2020 [“Reservations and Declarations for Treaty No.005 - Convention for the Protection of Human Rights and Fundamental Freedoms.” Treaty Office, Council of Europe.].

Some of Armenia's derogations have had a negative impact on vulnerable populations even though they were intended to be non-discriminatory, as per guidance from the UN’s Human Rights Office of the High Commissioner (OHCHR). Its decision to derogate from Article 9 of ICCPR was neither a necessary nor a proportionate measure to aid them in halting the public health crisis caused by the pandemic as it caused further issues to public health and to the judicial system.  One of the populations that was particularly affected by this derogation were detainees. Article 9 of the ICCPR which gives the right to all detainees to a fair and timely trial. By derogating from this article, the Armenian government disproportionately targeted detainees in comparison to the rest of the population. As the Armenian government grappled with the rapidly spreading pandemic they tried to slow its spread through the judicial system by halting court proceedings. This resulted in overcrowding of detaining facilities which posed a risk to the health of detained individuals and to public health [Mamulyan, Araks, and Narek Aleksanyan. “Armenia's Courts and the Coronavirus: Overcoming the Challenges.” Hetq, Hetq Investigative Journalism, 31 Dec. 2020, hetq.am/en/article/125964.]. The problem was exacerbated by the government’s decision to prolong its derogation multiple times.

Victims of domestic abuse were also unintendedly harmed by the government’s decision to proclaim the state of emergency and derogate from Article 21 of ICCPR and Article 11 of ECHR. As these proclamations meant that people could not leave their houses after a certain time and could not congregate with others, many women found themselves trapped in their homes under constant surveillance of their abusers. The police department recorded no increase in reports of domestic abuse, however, the Ombudsperson’s Office and various NGOs received an influx of reports after some measures had been lifted [Mkrtchyan, Gayane. “Armenia: Activists Call for More Action on Abuse.” Institute for War and Peace Reporting, 4 June 2020, iwpr.net/global-voices/armenia-activists-call-more-action-abuse.]

>>>>

---
title: "Derogations and Democratic Backsliding: Exploring the Pandemic's Effects on Civic Spaces"
author:
- name: Suparna Chaudhry
  affiliation: Lewis & Clark College
  email: schaudhry@lclark.edu
  url: http://www.suparnachaudhry.com/
- name: Andrew Heiss
  affiliation: Georgia State University
  email: aheiss@gsu.edu
  url: https://www.andrewheiss.com/
date: March 21, 2021
published: This is an early draft of a working paper.
git-repo: https://github.com/andrewheiss/covid-derogations
abstract: >-
  In an effort to combat the COVID-19 pandemic, most governments have imposed restrictions on civic freedoms in the interest of public health. In many cases, these emergency measures have been temporary and proportional in terms of controling the pandemic. In many other instances, however, governments have used these restrictions to suppress opposition and more permanently restrict civic space. Systematically measuring the consequences of COVID restrictions, however, is a difficult task. We examine two possible quantitative measures of the relationship between of COVID restrictions and civil society space. First, we use the Variety of Democracy project's newly released Pandemic Violations of Democratic Standards Index to explore if and how civil society restrictions predict pandemic backsliding. Second, while many countries sign international human rights treaties that ostensibly bind states to respect rights, several treaties allow for emergency derogations from these obligations. We tabulate formal human rights treaty derogations due to the pandemic and explore whether these emergency measures led to lasting declines in associational and human rights. We hope that our exploration of these two measures will provide rich data-based descriptions of the relationship between COVID restrictions and civic space that will allow for more causal work in the future.
reference-section-title: References
bibliography: bibliography.bib
csl: pandoc/csl/chicago-author-date.csl
output:
  bookdown::pdf_document2: 
    template: pandoc/templates/xelatex.tex
    dev: cairo_pdf
    pandoc_args: 
      - "--top-level-division=section"
      - "--shift-heading-level-by=0"
      - "-V"
      - "bibstyle-chicago-authordate"
      - "-V"
      - "chapterstyle=hikma-article"
    md_extensions: +raw_tex
    latex_engine: xelatex
    toc: no
    keep_tex: yes
    citation_package: biblatex
  bookdown::html_document2: 
    number_sections: false
    template: pandoc/templates/html.html
    css: pandoc/templates/ath-clean.css
    md_extensions: +raw_tex
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.retina = 3,
                      tidy.opts = list(width.cutoff = 120),  # For code
                      options(width = 90),  # For output
                      fig.asp = 0.618, fig.width = 7, 
                      fig.align = "center", out.width = "85%",
                      echo = FALSE, warning = FALSE, message = FALSE)

options(dplyr.summarise.inform = FALSE,
        knitr.kable.NA = "")

library(tidyverse)
library(knitr)
library(kableExtra)
library(scales)
library(janitor)
library(ggbeeswarm)
library(patchwork)
library(sf)
library(here)

source(here("lib", "graphics.R"))

# Generated via random.org
set.seed(9951)

pandem_clean <- readRDS(here("data", "derived_data", "pandem_clean.rds"))
pandem_single <- readRDS(here("data", "derived_data", "pandem_single.rds"))
derogations_clean <- readRDS(here("data", "derived_data", "derogations_clean.rds"))
vdem_clean <- readRDS(here("data", "derived_data", "vdem_clean.rds"))
pandem_derog <- readRDS(here("data", "derived_data", "pandem_derog.rds"))
civicus_clean <- readRDS(here("data", "derived_data", "civicus_clean.rds"))

world_map <- read_sf(here("data", "raw_data", "ne_110m_admin_0_countries",
                          "ne_110m_admin_0_countries.shp")) %>% 
  filter(ISO_A3 != "ATA")

matrix_from_vector <- function(x, ncol) {
  n_balanced <- ceiling(length(x) / ncol) * ncol
  matrix(c(x, rep(NA, n_balanced - length(x))), ncol = ncol)
}
```

```{r calc-derogations}
derog_count <- derogations_clean %>% 
  filter(ratified == 1) %>% 
  count(derogate)
```

<!-- manuscript.md is generated from manuscript.Rmd. Only edit the .Rmd file, *not* the .md file. -->

Combating a deadly virus like SARS-CoV-2 requires extraordinary public health measures that often conflict with personal rights and freedoms. In an effort to fight the COVID-19 pandemic, governments across the world have imposed restrictions on the freedom of movement, association, assembly, and other civic freedoms in the interest of public health. While countries are ostensibly free to make these emergency domestic policy decisions within their own borders, many states are also bound by international human rights treaties. Throughout the COVID-19 pandemic, many countries have declared states of emergency and formally derogated (i.e. temporarily suspended) their international treaty obligations. However, far more countries have adopted emergency measures and restricted individual liberties than have derogated from human rights treaties. While some states that have formally derogated with temporary and proportionate emergency responses, as advised by the United Nations Office of the High Commissioner of Human Rights [@ohchr:2020], others have used these emergency measures to suppress opposition and more permanently restrict civic space. During the COVID-19 pandemic, how many countries have formally derogated from their human rights treaty obligations and how has the nature of measures implemented during this period differed across countries? In the long-term, we aim to use our rich data-based description of the relationship between pandemic restrictions and civic space to examine whether these emergency measures led to lasting declines in associational and human rights, furthering the phenomenon of closing civic space.

```{r calc-civicus}
civicus <- civicus_clean %>%
  count(category) %>% 
  mutate(prop = n / sum(n)) %>% 
  mutate(prop_nice = percent_format(accuracy = 1)(prop)) %>% 
  split(.$category)
```

Studying state compliance with the international human rights treaty obligations as well as the (unintended) effects of derogations is important, especially given the ongoing prior phenomenon of closing civic space. Even prior to the pandemic, since 2013, 91 countries had proposed or enacted 244 measures restricting civil society [@icnl:2021]. In Human Rights Watch's 2016 World Report, Executive Director Kenneth Roth argued that civil society was under more aggressive attack than at any time in recent memory, and over the past decade, the majority of the world has seen a substantial narrowing of allowable civic space. As seen in Figure \@ref(fig:civicus-map), by the end of 2020, only `r civicus$Open$prop_nice` of countries had open and unrestricted civil societies [@civicusmonitor; @ChaudhryHeiss:2021].

```{r civicus-map, fig.cap="2020 CIVICUS Monitor civic space ratings"}
map_with_civicus <- world_map %>% 
  # Fix some Natural Earth ISO weirdness
  mutate(ISO_A3 = ifelse(ISO_A3 == "-99", as.character(ISO_A3_EH), as.character(ISO_A3))) %>% 
  mutate(ISO_A3 = case_when(
    .$ISO_A3 == "GRL" ~ "DNK",
    .$NAME == "Norway" ~ "NOR",
    .$NAME == "Kosovo" ~ "XKK",
    TRUE ~ ISO_A3
  )) %>% 
  left_join(civicus_clean, by = c("ISO_A3" = "iso3c"))

ggplot() +
  geom_sf(data = map_with_civicus, aes(fill = fct_rev(category)), size = 0.15, color = "black") +
  coord_sf(crs = st_crs("ESRI:54030")) +  # Robinson
  scale_fill_viridis_d(option = "plasma", end = 0.9, 
                       na.translate = FALSE, name = "Civic space") +
  theme_pandem_map() + 
  theme(legend.key.size = unit(0.7, "lines"))
```

According to the International Center for Not-for-Profit Law (ICNL), since the beginning of the pandemic in March 2020, 107 countries have declared official states of emergency, with 56 implementing measures affecting freedom of expression and an astonishing 139 imposing measures affecting freedom of assembly [@icnl-covid]. Many other countries have erected new barriers to public information, limited media access, engaged in government-sponsored disinformation campaigns, and penalized dissent. Some countries have prevented NGOs from delivering services and banned civil society organizations from carrying out their missions. Others have used pandemic restrictions as an excuse for delaying elections [@ifes:2020] or to denying arrested individuals adequate legal representation in judicial proceedings [@unodc]. The pandemic has led to a marked increase in human rights violations around the world.

States that are signatories to international human rights treaties are arguably bound to uphold them and limit potential violations of human rights. While many countries have been able to implement emergency public health measures without (1) requesting formal derogations from international treaties or (2) violating mass human rights, many others have essentially violated their treaty commitments. While issuing derogations is not necessarily a sign of impending human rights violations, several countries have already renewed their initial derogations, leading to concerns of a rise in serial derogators and a potential for emergency measures that are neither proportional nor temporary.

We explore the relationship between emergency pandemic responses, treaty derogations, human rights, democratiziation, and the space for civil society using original data on derogations from the International Covenant on Civil and Political Rights (ICCPR) during the pandemic as well as new data collected by the Varieties of Democracy (V-Dem) project on emergency measures implemented during the COVID-19 pandemic. After looking at at some of the predictors of the adoption of emergency measures, the violation of human rights, and the choice to derogate, we develop a typology of pandemic responses and find that most countries failed to request formal ICCPR derogations despite the adoption of strict publich health measures. Only `r filter(derog_count, derogate == 1)$n` countries submitted formal derogations as part of their pandemic response, and of these, five adopted policies that were not temporary, proportional, or necessary. We then provide short case studies of Guatemala, Armenia, Egypt, and India to illustrate different types of choices to derogate and declare emergencies. We conclude with brief thoughts about the potential of using this data for future work.

# International treaty derogations

Derogations in international law authorize states to temporarily suspend their international treaty commitments during times of crises, providing them with flexibility in responding to national crises. However, because derogations are only triggered during extraordinary times of war, disease, and other extenuating circumstances, these suspensions of obligations occur when individual liberties are under serious threat [@HBderogations, 674, 677]. By declaring a state of public emergency and formally derogating, states acknowledge that these measures are temporary, necessary, and proportional to the interest at stake, with an aim to restore normalcy as soon as possible.

Derogations—especially through the use of formal notifications by the state—provide vital information to both international and domestic monitoring bodies, interest groups, and advocates about which rights are suspended, for how long, and the reasoning behind these suspensions. This information allows these actors—at least in principle—to challenge measures that are excessive, vague, or outlast the intended time frame of their implementation [@Helfer:2021]. Derogations also provide an important temporary escape hatch—without the ability to derogate, countries facing emergencies may outrightly violate their treaty commitments under international law. Further, they may also illegally violate non-derogable rights such as right to life, freedom from torture, and slavery.

Derogations in international law authorize states to temporarily suspend their international treaty commitments during times of crises, providing them with flexibility in responding to national crises. However, because derogations are only triggered during extraordinary times of war, disease, and other extenuating circumstances, these suspensions of obligations occur when individual liberties are under serious threat [@HBderogations, 674, 677]. By declaring a state of public emergency and formally derogating, states acknowledge that these measures are temporary, necessary, and proportional to the interest at stake, with an aim to restore normalcy as soon as possible. Derogations provide an important temporary escape hatch—without the ability to derogate, countries facing emergencies may outrightly violate their treaty commitments under international law. Further, they may also illegally violate non-derogable rights such as right to life, freedom from torture, and slavery.

While the most frequent derogators are stable democracies and countries where domestic courts can exercise strong oversight of the executive and hold them responsible for breaches of human rights agreements, derogations—especially in the context of a long-lasting event such like a pandemic—are more concerning because of the tendency of some countries to turn into "serial derogators." These countries generally derogate "without providing information about rights restrictions and in multiple consecutive years" [@HBderogations, 675]. These patterns are typically encountered in countries where the judiciary is weak and voters cannot easily remove leaders from office.

@@

The exceptional circumstances brought on by the global pandemic have led to numerous extensive derogations. International human rights treaties typically lay out circumstances during which derogations may take place. For signatories of the European Convention on Human Rights (ECHR), for instance, derogations can be made "in time of war or other public emergency threatening the life of the nation" [@ohchr:2003, 815]; for the American Convention on Human Rights, derogations are allowed in "time of war, public danger, or other emergency that threatens the independence or security of a state" [@ohchr:2003, 859]. In the current version of this paper, however, we focus on derogations from—and violations of—the International Covenant on Civil and Political Rights (ICCPR), arguably the most significant and extensive treaties in the international human rights regime.

The ICCPR was adopted by the UN General Assembly on December 19, 1966 and came into force on March 23, 1976. It was designed to recognize "the inherent dignity and of the equal and inalienable rights of all members of the human family" and it outlined dozens of articles aimed at creating conditions where everyone can enjoy their civil, political, economic, social, and cultural rights. Like other human rights treaties, the ICCPR allows the suspension or derogation of certain civil and political rights only under specific situations "public emergency which threatens the life of the nation" [@iccpr].

Recognizing that public health exigencies would require states to curb civil and political rights, the UN Human Rights Office of the High Commissioner (OHCHR) published guidance on emergency measures on April 27 2020, stating that any emergency measures must meet specific criteria of legality, necessity, proportionality, and non-discrimination. Further, the OHCHR reminded countries that states of emergency are specifically regulated under human rights law and that "states should take measures to prevent human rights violations and abuses associated with the state of emergency perpetrated by state and non-state actors" [@ohchr:2020]. The OHCHR concluded with a reminder of the necessity of maintaining democratic norms and practices, including transparency, the right to information, media freedom, respect for rule of law, and allowing dissent. It is worth noting, however, that no international court or monitoring body has assessed whether or not individual countries' measures are necessary, temporary, proportionate.

However, given the ongoing trend of limiting civic space, many of these democratic freedoms have been under assault for years. Additionally, trends in democratic backsliding more generally raise concerns that many states may be using emergency measures intended to deal with the pandemic as a way to permanently suppress civic space and limit human rights. How many countries have formally derogated from their treaty obligations? How many have imposed official states of emergency? How have democratic norms, respect for human rights, and the environment for civil society fared under these emergency measures? Does formal derogation provide some protection against human rights abuses? Most importantly, will these emergency measures lead to prolonged abuses and reduced civic space?

# Derogations, emergencies, human rights, and civil society

We explore these questions with both quantitative and qualitative data. In 2020, the Varieties of Democracy (V-Dem) project published "PanDem," a new dataset on pandemic-era democratic backsliding [@vdem-pandem]. This PanDem data tracks several important measures of democratization and human rights abuses, including data on national emergency measures, violations of non-derogable rights, limitations on the media and journalists, and military involvement in public health enforcement. PanDem also provides two indices measuring general violations of democratic standards and democratic backsliding. PanDem's data reveals fascinating relationships between pandemic emergency measures, democratization, and human rights.

```{r calc-emergency-responses}
em_response <- pandem_single %>% 
  count(emlaw_cond_min) %>% 
  mutate(prop = n / sum(n)) %>% 
  mutate(prop = percent_format(accuracy = 1)(prop)) %>% 
  mutate(emlaw_cond_min = make_clean_names(emlaw_cond_min)) %>% 
  split(.$emlaw_cond_min)

em_response$countries <- pandem_single %>% 
  filter(emlaw_cond_min == "No emergency response") %>% 
  pull(country_name) %>% 
  knitr::combine_words()
```

## Emergency measures, human rights, and civil society

A majority of countries made some sort of official public health response to the pandemic, and most used legal avenues for their responses (see Table \@ref(tab:tbl-emergency-responses)): `r em_response$emergency_response_with_legal_instruments$prop` of countries used legal instruments to declare public emergencies at some point during 2020, while `r em_response$emergency_response_without_legal_instruments$prop` used extralegal avenues. Only a handful had no official national emergency response throughout 2020: `r em_response$countries`.

```{r tbl-emergency-responses}
pandem_single %>% 
  count(emlaw_fct_min, name = "N") %>% 
  mutate(`%` = N / sum(N)) %>% 
  mutate(`%` = percent_format(accuracy = 0.1)(`%`)) %>% 
  rename("Emergency response" = emlaw_fct_min) %>% 
  kbl(caption = "National emergency responses to the COVID-19 pandemic", 
      booktabs = TRUE, linesep = "") %>% 
  kable_styling(font_size = 9)
```

There are distinct patterns that help predict which kind of response a country made (see Figure \@ref(fig:democracy-regime-declaration)). There is a uniform distribution of adherence to democratic values within countries that used legal methods for declaring emergencies. For instance, both Denmark and Tajikistan used legal instruments to declare their respective emergencies. However, among countries that declared emergencies without any legal instruments, there are no countries that score above 0.5 on V-Dem's liberal democracy index. All countries that used extralegal methods for handling the pandemic have weak democratic norms. This is also true when looking specifically at regime type. Only one democracy—Sri Lanka—declared an emergency extralegally.

```{r democracy-regime-declaration, fig.width=9.5, fig.height=4, fig.asp=NULL, out.width="100%", fig.cap="Emergency pandemic responses and democracy"}
plot_democracy <- pandem_single %>% 
  ggplot(aes(y = fct_rev(emlaw_cond_min), x = v2x_libdem, fill = emlaw_cond_min)) +
  geom_quasirandom(groupOnX = FALSE, width = 0.2, 
                   size = 1.75, color = "white", pch = 21) +
  guides(fill = FALSE) +
  scale_fill_viridis_d(option = "plasma", end = 0.8) +
  scale_y_discrete(labels = function(x) str_wrap(x, width = 15)) +
  coord_cartesian(xlim = c(0, 1)) +
  labs(x = "V-Dem liberal democracy index", y = NULL,
       caption = "Lower values indicate less democracy",
       title = "Level of democracy") +
  theme_pandem(9) +
  theme(panel.grid.major.x = element_blank(),
        plot.title = element_text(size = rel(1)))

plot_regime <- pandem_single %>% 
  ggplot(aes(x = regime, y = fct_rev(emlaw_cond_min), fill = emlaw_cond_min)) +
  geom_point(position = position_jitter(width = 0.2, height = 0.2),
             size = 1.75, color = "white", pch = 21) +
  scale_fill_viridis_d(option = "plasma", end = 0.8) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 15)) +
  scale_y_discrete(labels = function(x) str_wrap(x, width = 15)) +
  guides(fill = FALSE) +
  labs(x = "V-Dem Regimes of the World index", y = NULL,
       title = "Regime type") +
  theme_pandem(9) +
  theme(panel.grid.major.x = element_blank(),
        plot.title = element_text(size = rel(1)))

plot_democracy + plot_regime
```

An overarching fear as countries declared emergency pandemic measures is that these policies would serve as an excuse to violate human rights, limit the space for civil society, and lead to democratic backsliding. PanDem includes data on the intensity of various human rights violations, including discriminatory actions (measured on a 0–3 scale), violations of non-derogable rights like the right to life, freedom from torture, and freedom of thought, conscience, and religion, and physical violence by the police or military (measured on a 0–3 scale). The variety of pandemic-justified human rights violations differs by countries' formal responses.

```{r violation-plots, fig.width=9.5, fig.height=3.25, fig.asp=NULL, out.width="100%", fig.cap="Human rights violations across different emergency responses"}
violation_summary <- pandem_single %>% 
  group_by(emlaw_cond_min) %>% 
  summarize(across(c(discrim, ndrights, pomviol),
                   ~mean(.)))

plot_violation_ndrights <- violation_summary %>% 
  pivot_longer(-emlaw_cond_min) %>% 
  filter(name == "ndrights") %>% 
  ggplot(aes(y = value, x = emlaw_cond_min, color = emlaw_cond_min)) +
  geom_pointrange(aes(ymin = 0, ymax = value)) +
  guides(color = FALSE) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 15)) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_color_viridis_d(option = "plasma", end = 0.8) +
  coord_cartesian(ylim = c(0, 0.15)) +
  labs(y = "Proportion violating non-derogable rights", x = NULL,
       title = "Violations of non-derogable rights") +
  theme_pandem(9) +
  theme(panel.grid.major.x = element_blank(),
        plot.title = element_text(size = rel(1)))

plot_violation_discrim <- violation_summary %>% 
  pivot_longer(-emlaw_cond_min) %>% 
  filter(name == "discrim") %>% 
  ggplot(aes(y = value, x = emlaw_cond_min, color = emlaw_cond_min)) +
  geom_pointrange(aes(ymin = 0, ymax = value)) +
  guides(color = FALSE) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 15)) +
  scale_color_viridis_d(option = "plasma", end = 0.8) +
  labs(y = "Average discriminatory measures", x = NULL,
       caption = "Original scale ranges from 0–3",
       title = "Discriminatory measures") +
  theme_pandem(9) +
  theme(panel.grid.major.x = element_blank(),
        plot.title = element_text(size = rel(1)))

plot_violation_pomviol <- violation_summary %>% 
  pivot_longer(-emlaw_cond_min) %>% 
  filter(name == "pomviol") %>% 
  ggplot(aes(y = value, x = emlaw_cond_min, color = emlaw_cond_min)) +
  geom_pointrange(aes(ymin = 0, ymax = value)) +
  guides(color = FALSE) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 15)) +
  scale_color_viridis_d(option = "plasma", end = 0.8) +
  labs(y = "Average abusive enforcement", x = NULL,
       caption = "Original scale ranges from 0–3",
       title = "Abusive enforcement") +
  theme_pandem(9) +
  theme(panel.grid.major.x = element_blank(),
        plot.title = element_text(size = rel(1)))

plot_violation_discrim + plot_violation_ndrights + plot_violation_pomviol
```

As seen in Figure \@ref(fig:violation-plots), countries that implemented emergency measures without a formal legal process were more likely to engage in discriminatory measures and violate non-derogable rights. Surprisingly, countries that used formal legal instruments were more likely to enforce these measures harshly and have a higher average abusive enforcement score. This is possibly due the the fact that these countries' states of emergency were formalized and thus had official enforcement measures—there were administrative and legislative teeth to these regulations, so to speak. Countries that used extralegal measures for their emergencies had fewer legal avenues to abuse when enforcing.

The level of democracy in a country is a fairly strong predictor of the severity of its pandemic response. As seen in Figure \@ref(fig:libdem-pandem), stronger democracies have substantially lower pandemic violations scores than countries with no democracy. Democracy is also a predictor of backsliding, though it has a curvilinear relationship due to the nature of backsliding. Strong democracies and non-democracies are at the lowest risk of backsliding already—strong consolidated democracies are ostensibly more resilient to authoritarian movements, while non-democracies have little democracy to undo. Countries with midrange democracy scores are at the highest risk of backsliding as a result of their emergency pandemic measures.

```{r libdem-pandem, message=FALSE, fig.width=9.5, fig.height=4, fig.asp=NULL, out.width="100%", fig.cap="Relationship between liberal democracy, pandemic violations, and pandemic backsliding"}
plot_libdem_pandem <- pandem_derog %>% 
  ggplot(aes(x = v2x_libdem, y = pandem)) +
  geom_point(aes(fill = emlaw_cond_min), pch = 21, color = "white", size = 1.75) +
  geom_smooth(method = "lm") +
  scale_fill_viridis_d(option = "plasma", end = 0.8) +
  guides(fill = guide_legend(ncol = 1)) +
  labs(x = "V-Dem liberal democracy index", y = "V-Dem pandemic violations index", fill = NULL) +
  theme_pandem(9)

plot_libdem_panback <- pandem_derog %>% 
  ggplot(aes(x = v2x_libdem, y = panback)) +
  geom_point(aes(fill = emlaw_cond_min), pch = 21, color = "white", size = 1.75) +
  geom_smooth(method = "gam") +
  scale_fill_viridis_d(option = "plasma", end = 0.8) +
  guides(fill = guide_legend(ncol = 1)) +
  labs(x = "V-Dem liberal democracy index", y = "V-Dem pandemic backsliding index", fill = NULL) +
  theme_pandem(9)

plot_libdem_pandem + plot_libdem_panback
```

## Derogations, emergency measures, and human rights

Countries differ in how they have aligned and justified their emergency pandemic measures with their international treaty obligations. V-Dem's PanDem project does not include data on whether countries formally derogated from treaties. Accordingly, we collected our own original dataset listing all formal derogations from the ICCPR, which we will later expand to other treaties like the ECHR, ACHR, and the International Covenant on Economic, Social and Cultural Rights (CESCR).

Of the `r sum(derog_count$n)` countries that have ratified the ICCPR, `r filter(derog_count, derogate == 1)$n` submitted formal derogations at some point during the pandemic. We combine our list of derogations with the PanDem data to determine how these derogations have respected human rights and resulted in measures that are temporary, proportional, or necessary. For the sake of this paper, we identify measures that are not temporary, proportional, or necessary based on whether (1) a declared emergency has an end date and is not repeated or renewed excessively, (2) discriminatory violations have been de jure minor, de jure major, or de facto major, (3) non-derogable rights have been violated, and (4) abusive enforcement has occurred often or is widespread. Identifying both "proper" and "improper" responses like this allows us to create a typology of derogations and emergency measures, summarized in Table \@ref(tab:tbl-derog-typology).[^typology-note] Table \@ref(tab:tbl-derog-a-b) lists all countries that formally derogated from ICCPR obligations, categorized by whether their emergency measures were temporary, proportional or necessary. In the appendix we include a list of countries that did not formally derogate (see Table \@ref(tab:tbl-derog-c-d)).

[^typology-note]: PanDem only includes data for `r nrow(pandem_single)` countries, so the total here does not equal the ICCPR's `r sum(derog_count$n)` ratifying states.
